Collaborative Learning with Dirichlet Process Clustering for Rapid Online Adaptation in Robotics

### 2023-08-28 paper/paper.tex
Aug 28

\begin{algorithm}
\caption{MAP-Elites}
\begin{algorithmic}
\STATE \textbf{procedure} MAP-Elites
\STATE Discretize the task space into grids
\STATE $\mathcal{R} \leftarrow \emptyset$ \COMMENT{Initialize repertoire}
\FOR{$\bm{i} = 1 \rightarrow G$}
\STATE $\theta \leftarrow $ generate{\_}random{\_}policy()
\STATE $\Theta$, $f$ $\leftarrow$ evaluate{\_}policy($\theta$) \COMMENT{Get outcome and fitness}
\STATE add{\_}to{\_}repertoire($\Theta$, $f$, $\theta$, $\mathcal{R}$) \COMMENT{Update repertoire}
\ENDFOR
\FOR{$\bm{i} = 1 \rightarrow T$}
\STATE $\vartheta \leftarrow$ select{\_}parents($\mathcal{R}$)
\STATE $\bm{\theta} \leftarrow$ crossover($\vartheta$, $N$) \COMMENT{Generate $N$ off-springs}
\FOR{$\bm{j} = 1 \rightarrow N$}
\STATE $\Theta$, $f$ $\leftarrow$ evaluate{\_}policy($\bm{\theta_j}$)
\COMMENT{Off-spring evaluation}
\STATE add{\_}to{\_}repertoire($\Theta$, $f$, $\bm{\theta_j}$, $\mathcal{R}$)
\ENDFOR
\ENDFOR
\end{algorithmic}
\end{algorithm}
\noindent
Evaluating the policies in the simulation is the major source of computationally expense for the MAP-Elites.
A merit of this algorithm is that the inner for loop of off-spring evaluation can be conducted in parallel, and the repertoire update can be made after having collected all the results of the off-springs.
This enables us to generate a large number of off-springs in each iteration and evaluate them in parallel, fully leveraging the power of modern Cluster computing \cite{Q-Dax}.
By running the algorithm for a few thousand iterations, we will end up having a repertoire of high-performing policies well-covering the task space.

@@

\subsection{Online Learning the Dynamics}
After having clustered the historical data, we can use the result to assist adaptation.
In RTE, we learn the distortion online using GPs, where the parameters (prior mean function, kernel etc.) are determined manually.
In our method, the dynamics is modelled with an infinite mixture of GPs. 
The GP parameters are given by the posterior distribution found with historical data, which is a mixture model of finite terms:
\begin{equation}
p(\bm{\theta}) = \sum_j \frac{n_j}{n + \alpha} p(\bm{\theta}|c=j) 
+ \frac{\alpha}{n + \alpha} H(\bm{\theta})
\label{trained_mixture_model}
\end{equation}
where the summation is made over the clusters containing at least one episode, and $p(\bm{\theta}|c=j)$ denotes the posterior for the $j^{\text{th}}$ cluster.
Each instance of $\bm{\theta}$ sampled from $p(\bm{\theta})$ is a certain configuration of parameters of GP.
The last term results from the important property of DPMM that there is always a non-zero probability that we will get a new cluster (face a new situation).
The online learned distortion is integrated over the posterior of $\bm{\theta}$ conditioned on the collected real-world interactions:
\begin{equation}
\begin{gathered}
p(x^*_d|\bm{x}_{1:N}) 
\\
= \sum_j p(c=j| \bm{x}_{1:N}) \int p(x^*_d|\bm{\theta}, \bm{x}_{1:N})p(\bm{\theta}|c=j, \bm{x}_{1:N}) d\bm{\theta} 
\\
+ p(c \neq j, \forall j|\bm{x}_{1:N}) \int p(x^*_d|\bm{\theta}, \bm{x}_{1:N}) p(\bm{\theta} | \bm{x}_{1:N}) d\bm{\theta}
\end{gathered}
\label{MGP_posterior}
\end{equation}
where the indicator probability is given by:
\begin{equation}
\begin{gathered}
p(c=j| \bm{x}_{1:N}) 
= \frac{n_j \cdot p(\bm{x}_{1:N}|c=j)
}{Z}
\\ 
p(c \neq j, \forall j| \bm{x}_{1:N})
= \frac{\alpha \cdot p(\bm{x}_{1:N}|c \neq j, \forall j)
}{Z}
\end{gathered}
\label{MGP_indicator}
\end{equation}
where $\bm{x}_{1:N}$ stands for the collected ($N$ number of) interaction data during deployment.
This is complicated and requires calculating multiple integrals over the posteriors. 
Fortunately, if we have applied posterior convergence estimate for $p(\bm{\theta}|c=j)$, we will have:
\begin{equation}
\begin{gathered}
\int p(x^*_d|\bm{\theta}, \bm{x}_{1:N})p(\bm{\theta}|c=j, \bm{x}_{1:N}) d\bm{\theta} \approx p(x^*_d|\bm{\theta}^*_j, \bm{x}_{1:N})
\\
\text{and } p(\bm{x}_{1:N}|c=j) \approx p(\bm{x}_{1:N}|\bm{\theta}^*_j)
\end{gathered}
\label{posterior_convergence_simplification}
\end{equation}
where $\bm{\theta}^*_j$ denotes the optimal parameters for the cluster $j$.
In actual implementation, the clustering of historical data with DP is conducted in the archive management system where expensive data processing is available. 
The results are saved as an adaptation system to be used (downloaded) by individual robots.
The adaptation system is independent from the archive, and we are free to make simplifications for embedded systems.
Hence, we can apply posterior convergence estimate for all the clusters we have found for simplicity.
For clusters that are not large enough to refit the parameters, we will just use the parameters of the most likely prior sample.
This might be inappropriate for clusters that are very small, and hence we can discard them from the adaptation system.

We still need to solve the two integrals in Eq. (\ref{MGP_posterior}) and Eq. (\ref{MGP_indicator}) for the case of a new cluster.
This can be hard even with the Monte Carlo sampling. 
If our MC prior does not cover this new situation, our method could fail.
In this paper, we take a very simple solution by using RTE to incorporate this new situation.
Although RTE is not optimized for any environment, it offers a satisfying baseline performance in all situations.
The non-zero probability of encountering new cases secured by DP and the use of RTE ensures that we can always adapt well to new situations and hence will never over-fit to the training data.
Combined with Eq. (\ref{posterior_convergence_simplification}), the distortion becomes:
\begin{equation}
\begin{gathered}
p(x^*_d|\bm{x}_{1:N}) 
= \sum_j p(c=j| \bm{x}_{1:N}) p(x^*_d|\bm{\theta}^*_j, \bm{x}_{1:N})
\\
+ \,\, p(c \neq j, \forall j|\bm{x}_{1:N})  p(x^*_d|\bm{x}_{1:N}, \text{RTE})
\end{gathered}
\label{simplified_distortion}
\end{equation}
where the indicator probabilities are:
\begin{equation}
\begin{gathered}
p(c=j| \bm{x}_{1:N}) 
= \frac{n_j \cdot p(\bm{x}_{1:N}|\bm{\theta}^*_j)
}{Z}
\\ 
p(c \neq j, \forall j| \bm{x}_{1:N})
= \frac{\alpha \cdot p(\bm{x}_{1:N}|\text{RTE})
}{Z}
\\ \text{where }
Z = \sum_k n_k \cdot p(\bm{x}_{1:N}|\bm{\theta}^*_k) + \alpha \cdot p(\bm{x}_{1:N}|\text{RTE})
\end{gathered}
\label{simplified_indicator}
\end{equation}
In contrast to RTE, our online learning of the dynamics consists of two levels.
The first level is to use the real-world data to figure out the situation that the robot is in, which is the indicator probability given by Eq. (\ref{simplified_indicator}).
Considering the dynamics can be distinct for different environments, the robot should be able to quickly figure out its current situation.
Since each situation has a prior mean function that can give estimation of the dynamics without further data, our method should be able to provide fast online learning of the real-world dynamics with much higher data-efficiency than RTE.
The second level is to learn the residual dynamics, which is the part of the distortion failed to be captured by the prior mean, using the kernels optimized for each situation.
The final prediction of dynamics given by Eq. (\ref{simplified_distortion}) is a mixture of Gaussians.
Where each mixture component has mean and variance calculated using Eq. (\ref{GP_posterior}), and the mixture weights are equal to the probabilities of being in each situation.

@@

\subsection{Robot Setup}

\subsection{Generating the Repertoire with MAP-Elites}

@@

\subsection{Online Learning the Dynamics}

>>>>

\begin{algorithm}
\caption{MAP-Elites}
\begin{algorithmic}
\STATE \textbf{procedure} MAP-Elites
\STATE Discretize the task space into grids
\STATE $\mathcal{R} \leftarrow \emptyset$ \COMMENT{Initialize repertoire}
\FOR{$\bm{i} = 1 \rightarrow G$}
\STATE $\theta \leftarrow $ generate{\_}random{\_}policy()
\STATE $\Theta$, $f$ $\leftarrow$ evaluate{\_}policy($\theta$) \COMMENT{Get outcome and fitness}
\STATE add{\_}to{\_}repertoire($\Theta$, $f$, $\theta$, $\mathcal{R}$) \COMMENT{Update repertoire}
\ENDFOR
\FOR{$\bm{i} = 1 \rightarrow T$}
\STATE $\vartheta \leftarrow$ select{\_}parents($\mathcal{R}$)
\STATE $\bm{\theta} \leftarrow$ crossover($\vartheta$, $N$) \COMMENT{Generate $N$ off-springs}
\FOR{$\bm{j} = 1 \rightarrow N$}
\STATE $\Theta$, $f$ $\leftarrow$ evaluate{\_}policy($\bm{\theta_j}$)
\COMMENT{Off-spring evaluation}
\STATE add{\_}to{\_}repertoire($\Theta$, $f$, $\bm{\theta_j}$, $\mathcal{R}$)
\ENDFOR
\ENDFOR
\end{algorithmic}
\label{MAP-Elites}
\end{algorithm}
\noindent
Evaluating the policies in the simulation is the major source of computationally expense for the MAP-Elites.
A merit of this algorithm is that the inner for loop of off-spring evaluation can be conducted in parallel, and the repertoire update can be made after having collected all the results of the off-springs.
This enables us to generate a large number of off-springs in each iteration and evaluate them in parallel, fully leveraging the power of modern Cluster computing \cite{Q-Dax}.
By running the algorithm for a few thousand iterations, we will end up having a repertoire of high-performing policies well-covering the task space.

@@

\subsection{Learning the Dynamics Online}
After having clustered the historical data, we can use the result to assist adaptation.
In RTE, we learn the distortion online using GPs, where the parameters (prior mean function, kernel etc.) are determined manually.
In our method, the dynamics is modelled with an infinite mixture of GPs. 
The GP parameters are given by the posterior distribution found with historical data, which is a mixture model of finite terms:
\begin{equation}
p(\bm{\theta}) = \sum_j \frac{n_j}{n + \alpha} p(\bm{\theta}|c=j) 
+ \frac{\alpha}{n + \alpha} H(\bm{\theta})
\label{trained_mixture_model}
\end{equation}
where the summation is made over the clusters containing at least one episode, and $p(\bm{\theta}|c=j)$ denotes the posterior for the $j^{\text{th}}$ cluster.
Each instance of $\bm{\theta}$ sampled from $p(\bm{\theta})$ is a certain configuration of parameters of GP.
The last term results from the important property of DPMM that there is always a non-zero probability that we will get a new cluster (face a new situation).
The online learned distortion is integrated over the posterior of $\bm{\theta}$ conditioned on the collected real-world interactions:
\begin{equation}
\begin{gathered}
p(x^*_d|\bm{x}_{1:N}) 
\\
= \sum_j p(c=j| \bm{x}_{1:N}) \int p(x^*_d|\bm{\theta}, \bm{x}_{1:N})p(\bm{\theta}|c=j, \bm{x}_{1:N}) d\bm{\theta} 
\\
+ p(c \neq j, \forall j|\bm{x}_{1:N}) \int p(x^*_d|\bm{\theta}, \bm{x}_{1:N}) p(\bm{\theta} | \bm{x}_{1:N}) d\bm{\theta}
\end{gathered}
\label{MGP_posterior}
\end{equation}
where the indicator probability is given by:
\begin{equation}
\begin{gathered}
p(c=j| \bm{x}_{1:N}) 
= \frac{n_j \cdot p(\bm{x}_{1:N}|c=j)
}{Z}
\\ 
p(c \neq j, \forall j| \bm{x}_{1:N})
= \frac{\alpha \cdot p(\bm{x}_{1:N}|c \neq j, \forall j)
}{Z}
\end{gathered}
\label{MGP_indicator}
\end{equation}
where $\bm{x}_{1:N}$ stands for the collected ($N$ number of) interaction data during deployment.
This is complicated and requires calculating multiple integrals over the posteriors. 
Fortunately, if we have applied posterior convergence estimate for $p(\bm{\theta}|c=j)$, we will have:
\begin{equation}
\begin{gathered}
\int p(x^*_d|\bm{\theta}, \bm{x}_{1:N})p(\bm{\theta}|c=j, \bm{x}_{1:N}) d\bm{\theta} \approx p(x^*_d|\bm{\theta}^*_j, \bm{x}_{1:N})
\\
\text{and } p(\bm{x}_{1:N}|c=j) \approx p(\bm{x}_{1:N}|\bm{\theta}^*_j)
\end{gathered}
\label{posterior_convergence_simplification}
\end{equation}
where $\bm{\theta}^*_j$ denotes the optimal parameters for the cluster $j$.
In actual implementation, the clustering of historical data with DP is conducted in the archive management system where expensive data processing is available. 
The results are saved as an adaptation system to be used (downloaded) by individual robots.
The adaptation system is independent from the archive, and we are free to make simplifications for embedded systems.
Hence, we can apply posterior convergence estimate for all the clusters we have found for simplicity.
For clusters that are not large enough to refit the parameters, we will just use the parameters of the most likely prior sample.
This might be inappropriate for clusters that are very small, and hence we can discard them from the adaptation system.

We still need to solve the two integrals in Eq. (\ref{MGP_posterior}) and Eq. (\ref{MGP_indicator}) for the case of a new cluster.
This can be hard even with the Monte Carlo sampling. 
If our MC prior does not cover this new situation, our method could fail.
In this paper, we take a very simple solution by using RTE to incorporate this new situation.
Although RTE is not optimized for any environment, it offers a satisfying baseline performance in all situations.
The non-zero probability of encountering new cases secured by DP and the use of RTE ensures that we can always adapt well to new situations and hence will never over-fit to the training data.
Combined with Eq. (\ref{posterior_convergence_simplification}), the distortion becomes:
\begin{equation}
\begin{gathered}
p(x^*_d|\bm{x}_{1:N}) 
= \sum_j p(c=j| \bm{x}_{1:N}) p(x^*_d|\bm{\theta}^*_j, \bm{x}_{1:N})
\\
+ \,\, p(c \neq j, \forall j|\bm{x}_{1:N})  p(x^*_d|\bm{x}_{1:N}, \text{RTE})
\end{gathered}
\label{simplified_distortion}
\end{equation}
where the indicator probabilities are:
\begin{equation}
\begin{gathered}
p(c=j| \bm{x}_{1:N}) 
= \frac{n_j \cdot p(\bm{x}_{1:N}|\bm{\theta}^*_j)
}{Z}
\\ 
p(c \neq j, \forall j| \bm{x}_{1:N})
= \frac{\alpha \cdot p(\bm{x}_{1:N}|\text{RTE})
}{Z}
\\ \text{where }
Z = \sum_k n_k \cdot p(\bm{x}_{1:N}|\bm{\theta}^*_k) + \alpha \cdot p(\bm{x}_{1:N}|\text{RTE})
\end{gathered}
\label{simplified_indicator}
\end{equation}
In contrast to RTE, our online learning of the dynamics consists of two levels.
The first level is to use the real-world data to figure out the situation that the robot is in, which is the indicator probability given by Eq. (\ref{simplified_indicator}).
Considering the dynamics can be distinct for different environments, the robot should be able to quickly figure out its current situation.
Since each situation has a prior mean function that can give estimation of the dynamics without further data, our method should be able to provide fast online learning of the real-world dynamics with much higher data-efficiency than RTE.
The second level is to learn the residual dynamics, which is the part of the distortion failed to be captured by the prior mean, using the kernels optimized for each situation.
The final prediction of dynamics given by Eq. (\ref{simplified_distortion}) is a mixture of Gaussians.
Where each mixture component has mean and variance calculated using Eq. (\ref{GP_posterior}), and the mixture weights are equal to the probabilities of being in each situation.
This learned dynamics can then be used by the robot to correct its motion accordingly.

@@

\subsection{Robot Setup}
To evaluate our methodology, experiment is conducted on the Ant robot in PyBullet simulation \cite{PyBullet} (see Fig \ref{Ant_robot}). 
This robot has 8 degrees of freedom (2 for each leg) corresponding to 8 motors powering its motion.
Note that the joints connecting the body and each leg can only rotate horizontally, meaning that the four thighs (the part of each leg directly attached to the body) are always on the same plane.
Hence, the Ant robot cannot lift up its leg to walk like real animals.
The motors employ PID positional control, where the torque is given by:
\begin{equation}
\tau = K_p \theta_e + K_i \int \theta_e(t) dt + K_d \frac{d \theta_e}{dt}
\label{PID}
\end{equation}
where $\theta_e = \theta_t - \theta$ ($\theta_t$ is the positional target).
In PyBullet, the maximum torque is $\pm 1$. 
For the motors on the body, $K_p$ is set to be 1.43 and $K_d$ is set to be 0.072.
For the motors on the knees, $K_p$ is 1.637 and $K_d$ is 0.082.
For all motors, $K_i$ is set to zero for simplicity. 
\begin{figure}[h]
\centering
\includegraphics[width=0.45\textwidth]{intact_robot.pdf}
\caption{The Ant robot in PyBullet simulation. 
The four motors mounted on the body can rotate from $-40^{\circ}$ to $40^{\circ}$. The knee flexion angle for each leg ranges from 
$30^{\circ}$ to $100^{\circ}$.
}
\label{Ant_robot}
\end{figure}

Like in RTE, we use a periodic controller for the robot.
The period is set to 1 second, which corresponds to 100 time-steps in simulation.
The targets within each period is parametrized in the same way as in \cite{cully2015robots}.
The target of each motor is a smoothed squared wave (see Fig \ref{targets}) governed by three parameters: amplitude $\alpha$, phase $\phi$ and duty cycle $\tau$.
All the three parameters range from 0 to 1.
The duty cycle stands for the proportion of time in one period that the squared wave is in its high state.
This squared wave is then smoothed by a Gaussian kernel to remove rapid changes.
In each time-step of the simulation, we give each motor the corresponding positional target, which will then be converted into torques via PID.
Thus, each controller is parametrized by a vector of 24 dimensions, which is hence defined to be the genotype of this controller.
%
%
\begin{figure}[H]
\centering
\includegraphics[width=0.45\textwidth]{targets.pdf}
\caption{The periodic motor targets corresponding to $\alpha = 0.8$, $\phi=100$ ms and $\tau=0.6$.
The squared wave is smoothed by a Gaussian kernel with $\sigma = 100$ ms.
}
\label{targets}
\end{figure}

In this paper, we consider three sources of distortions: friction, payload and damages to the legs of the robot.
To collect the distortions as the basis for the prior mean function, we need to model these sources in simulation.
Different frictions can be easily achieved by changing the simulation settings.
Leg damages are modelled by giving a fixed positional target to the motor mounted on the knee.
This fixed target overwrites the command of the controller, so that the damaged leg is always in the upper position (see Fig \ref{damaged_robot}).
However, for unknown reasons, changing the mass of the robot do not affect the dynamics in simulation. 
Hence, we take an alternative approach by discounting the torque to model the payload.
%
%
\begin{figure}[h]
\centering
\includegraphics[width=0.45\textwidth]{damaged_robot.pdf}
\caption{The robot walking with one leg (left blue-yellow leg) damaged.
The knee flexion angle of this leg is fixed at $30^{\circ}$ to reduce contact with the ground.
}
\label{damaged_robot}
\end{figure}

\subsection{Generating the Repertoire with MAP-Elites}
To implement MAP-Elites, the task space for our robot is defined to be its 2D displacement, spanning from -4.8 to 4.8 on both the $x$ and $y$ axes. 
This continuous space is then discretized into a $32 \times 32$ grid, yielding a resolution of 0.3 unit per cell.
To evaluate each genotype, we first decode it into the periodic targets and then execute them for 4 seconds.
The fitness is defined to be a reward of the lifespan (number of time-steps before the body of the robot hits the ground) penalized by the electrical cost \cite{QDgym}.
In each iteration of MAP-Elites, we generate 128 off-springs from 8 selected parents.

To design a crossover operation that balances exploitation and exploration,

The crossover operation should balance the exploitation and exploration by ensuring some inheritance of the parents to preserve good genes as well as some random variations for promoting diversity.

computation 10 cpu cores.

For each controller, apart from the 2D displacement, we also need to record the three parameters for arc-based dynamics representation calculated using Eq. (\ref{all_three_parameters}).

@@

\subsection{Dynamics Learning}
